Equivalence Testing and the Second Generation P-Value

### 2018-07-30 sgpv_vs_TOST.Rmd
updated version

The second generation *p*-value (SGPV) is a new descriptive statistic that was recently proposed to "improve rigor, reproducibility and transparency across science" (Blume, McGowan, Dupont, & Greevy, (2018). The SGPV is 'the proportion of data-supported hypotheses that are also null hypotheses'. The researcher specify an equivalence range around the null hypothesis that specifies values that are considered practically equivalent to the null-hypothesis. The SGPV is the proportion of the  confidence interval (CI) around the observed effect estimate that falls within this equivalence range. If the CI falls completely inside the equivalence range the SGPV is 1, and if the CI falls completely outside the equivalence rnge the SGPV is 0.

The SGPV has strong similarities with an already existing approach known as equivalence testing (Lakens, 2017; Rogers, Howard, & Vessey, 1993). In an equivalence test, an equivalence range is based upon a smallest effect size of interest. In the Two One-Sided Tests (TOST) approach to equivalence testing data is tested against the upper and lower bounds of the equivalence range (e.g., a difference of -2 and +2). If both one-sided tests reject the presence of effects more extreme than the equivalence bounds when can act as if there is no meaningful effect. Because two one-sided tests are performed, equivalence can be declared when a 1-2*alpha CI (e.g., when the alpha level is 0.05, a 90% CI)  falls completely within the equivalence range of -2 and +2.

@@

One the other hand, if the observed mean is 140, the equivalence test is not significant (the observed mean is far outside the equivalence range of 143 to 147), *t*(`r tost_res2$TOST_df`) = `r tost_res2$TOST_t1`, *p* = `r tost_res2$TOST_p1` (or more accuratelty, *p* > .999 as *p*-values are bounded between 0 and 1). Because the 95% CI falls completely outside the equivalence bounds, the SGPV is `r SGPV2` (see Figure 1).

@@

We can examine this situation be calculating the SGPV and performing the TOST for a situation where sample sizes are small and the equivalence range is narrow, such that the CI is more than twice as large as the equivalence range. 
```{r, include=FALSE}
step = 0.01

@@

We can again plot the two statistics against each other so see where they are unrelated (indicated by staight lines in the curve). We see the SGPV is 0.5 for a range of observed means where the p-value from the equivalence test still varies. It should be noted that in these calculations the *p*-values for the TOST procedure are *never* smaller than 0.05 (i.e., they do not get below 0.05 on the y-axis). In other words, we can't conclude equivalence based on any of the observed means. How is this possible? Remember that the TOST procedure consists of two one-sided tests against the upper and lower equivalence bound. The TOST *p*-value is smaller than 0.05 if the 90% CI falls completely between the upper and lower equivalence bounds. However, we are examining a scenario where the 90% CI is so wide that it never falls completely within the two equivalence bounds. As Lakens (2017) notes: “in small samples (where CIs are wide), a study might have no statistical power (i.e., the CI will always be so wide that it is necessarily wider than the equivalence bounds).” None of the *p*-values based on the TOST procedure are below 0.05, and thus, in the long run we have 0% power.
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 7*: The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure 6.

There is one last situation where the *p*-value from the TOST procedure and the SGPV are unrelated. This is when the CI overlaps with the upper and lower equivalence bound, but the CI is *not* twice as wide as the equivalence range. This fourth category exists because of the decision by Blume and collagues to set the SGPV to 0.5 whenever the CI is twice as wide as the equivalence range and the CI overlaps with both equivalence bounds. This means that there are situations where the CI interval overlaps with both equivalence bounds, while the CI is less than twice as large as the equivalence bound. For example, in the example below, we see an example where the CI is only 1.79 times as wide as the equivalence bounds, but the CI overlaps with the lower and upper equivalence bounds.
```{r, include=FALSE}
m <- 0.2
n <- 10
sd <- 1
d.lo <- -0.4
d.hi <- 0.4

@@

If the observed mean would be somewhat closer to 0, or further away from 0, the SGPV would remain constant (the CI width does not change, it completely overlaps with the equivalence range) while the *p*-value for the TOST procedure varies. We can see this in Figure 9 below. The SGPV is not set to 0.5, but is slightly higher than 0.5 across a range of means. How high the SGPV will be when the CI overlaps with the lower and upper equivalence bounds, but the CI is not twice as large as the equivalence range, depends on the width of the CI and the equivalence range. 
```{r, include=FALSE}
step = 0.01

@@

If we once more plot the two statistics against each other so see where they are unrelated (indicated by staight lines in the curve). We see the SGPV is 0.56 for a range of observed means where the *p*-value from the equivalence test still varies.
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 10*: The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure 6.

To conclude this section, there are four situations where the *p*-value from the TOST procedure is unrelated to the SGPV, In all these situation the *p*-value for the equivalence test differentiates tests with different means, but the SGPV does not. Therefore, as a purely descriptive statistic, the SGPV is more limited than the *p*-value from the TOST procedure. The proportion of overlap can be the same value when the observed mean is 0 or when the observed mean falls just inside the equivalence bound, and additional information (e.g., the 95% CI) is required to differentiate these situations. One way to mitigate this limitation of the SGPV would be to set the SGPV to 0.5 whenever the CI overlaps with both the upper and lower equivalence bound (irrespective of the width of the CI).

@@

It seems Blume et al (2018) where not aware of the existence of equivalence tests, and we believe that our explanation of the similarities between the TOST procedure and the SGPV provides some useful context to interpret the contribution of second generation *p*-values to the statistical toolbox. The novelty lies in its use as a descriptive statistic. The added benefit of calculating the proportion of overlap of a 95% CI with the equivalence range, and using this percentage to describe the data, remains somewhat unclear for practical puroposes. Nevertheless, our only goal is to clarify the relationship between a newly proposed statistic and the already existing TOST approach used to test for equivalence, and let researchers make an informed decision about which statistical approach provides the best answer to their question.

@@

Blume, J. D., McGowan, L. D., Dupont, W. D., & Greevy, R. A. (2018). Second-generation *p*-values: Improved rigor, reproducibility, & transparency in statistical analyses. PLOS ONE, 13(3), e0188299. https://doi.org/10.1371/journal.pone.0188299
Lakens, D. (2017). Equivalence Tests: A Practical Primer for t Tests, Correlations, and Meta-Analyses. Social Psychological and Personality Science, 8(4), 355–362. https://doi.org/10.1177/1948550617697177
Rogers, J. L., Howard, K. I., & Vessey, J. T. (1993). Using significance tests to evaluate equivalence between two experimental groups. Psychological Bulletin, 113(3), 553–565. http://dx.doi.org/10.1037/0033-2909.113.3.553

>>>>

The second generation *p*-value (SGPV) is a new descriptive statistic that was recently proposed to "improve rigor, reproducibility and transparency across science" (Blume, McGowan, Dupont, & Greevy, (2018). The SGPV is 'the proportion of data-supported hypotheses that are also null hypotheses'. The researcher specify an equivalence range around the null hypothesis that specifies values that are considered practically equivalent to the null-hypothesis. The SGPV is the proportion of the  confidence interval (CI) around the observed effect estimate that falls within this equivalence range. If the CI falls completely inside the equivalence range the SGPV is 1, and if the CI falls completely outside the equivalence range the SGPV is 0.

The SGPV has strong similarities with an already existing approach known as equivalence testing (Lakens, 2017; Rogers, Howard, & Vessey, 1993). In an equivalence test, an equivalence range is based upon a smallest effect size of interest. In the Two One-Sided Tests (TOST) approach to equivalence testing data is tested against the upper and lower bounds of the equivalence range (e.g., a difference of -2 and +2). If both one-sided tests reject the presence of effects more extreme than the equivalence bounds we can act as if there is no meaningful effect. Because two one-sided tests are performed, equivalence can be declared when a 1-2*alpha CI (e.g., when the alpha level is 0.05, a 90% CI)  falls completely within the equivalence range of -2 and +2.

@@

One the other hand, if the observed mean is 140, the equivalence test is not significant (the observed mean is far outside the equivalence range of 143 to 147), *t*(`r tost_res2$TOST_df`) = `r tost_res2$TOST_t1`, *p* = `r tost_res2$TOST_p1` (or more accurately, *p* > .999 as *p*-values are bounded between 0 and 1). Because the 95% CI falls completely outside the equivalence bounds, the SGPV is `r SGPV2` (see Figure 1).

@@

We can examine this situation by calculating the SGPV and performing the TOST for a situation where sample sizes are small and the equivalence range is narrow, such that the CI is more than twice as large as the equivalence range. 
```{r, include=FALSE}
step = 0.01

@@

We can again plot the two statistics against each other so see where they are unrelated (indicated by straight lines in the curve). We see the SGPV is 0.5 for a range of observed means where the p-value from the equivalence test still varies. It should be noted that in these calculations the *p*-values for the TOST procedure are *never* smaller than 0.05 (i.e., they do not get below 0.05 on the y-axis). In other words, we can't conclude equivalence based on any of the observed means. How is this possible? Remember that the TOST procedure consists of two one-sided tests against the upper and lower equivalence bound. The TOST *p*-value is smaller than 0.05 if the 90% CI falls completely between the upper and lower equivalence bounds. However, we are examining a scenario where the 90% CI is so wide that it never falls completely within the two equivalence bounds. As Lakens (2017) notes: “in small samples (where CIs are wide), a study might have no statistical power (i.e., the CI will always be so wide that it is necessarily wider than the equivalence bounds).” None of the *p*-values based on the TOST procedure are below 0.05, and thus, in the long run we have 0% power.
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 7*: The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure 6.

There is one last situation where the *p*-value from the TOST procedure and the SGPV are unrelated. This is when the CI is wider than the equivalence range (so the precision is low) and overlaps with the upper and lower equivalence bound, but the CI is *not* twice as wide as the equivalence range. This fourth category exists because of the decision by Blume and colleagues to set the SGPV to 0.5 whenever the CI is twice as wide as the equivalence range and the CI overlaps with both equivalence bounds. This means that there are situations where the CI interval overlaps with both equivalence bounds, while the CI is less than twice as large as the equivalence bound. For example, in the example below, we see that the CI is only 1.79 times as wide as the equivalence bounds, but the CI overlaps with the lower and upper equivalence bounds.
```{r, include=FALSE}
m <- 0.2
n <- 10
sd <- 1
d.lo <- -0.4
d.hi <- 0.4

@@

If the observed mean would be somewhat closer to 0, or further away from 0, the SGPV would remain constant (the CI width does not change, it completely overlaps with the equivalence range) while the *p*-value for the TOST procedure can vary between 0 and .025. We can see this in Figure 9 below. The SGPV is not set to 0.5, but is slightly higher than 0.5 across a range of means. How high the SGPV will be when the CI overlaps with the lower and upper equivalence bounds, but the CI is not twice as large as the equivalence range, depends on the width of the CI and the equivalence range. 
```{r, include=FALSE}
step = 0.01

@@

If we once more plot the two statistics against each other so see where they are unrelated (indicated by straight lines in the curve). We see the SGPV is 0.56 for a range of observed means where the *p*-value from the equivalence test still varies.
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 10*: The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure 9.

To conclude this section, there are four situations where the *p*-value from the TOST procedure is unrelated to the SGPV. In all these situations the *p*-value for the equivalence test differentiates tests with different means, but the SGPV does not. Therefore, as a purely descriptive statistic, the SGPV is more limited than the value from the TOST procedure. The proportion of overlap can be the same value when the observed mean is 0 or when the observed mean falls just inside the equivalence bound, and additional information (e.g., the 95% CI) is required to differentiate these situations. One way to mitigate this limitation of the SGPV would be to set the SGPV to 0.5 whenever the CI overlaps with both the upper and lower equivalence bound (irrespective of the width of the CI).

@@

It seems Blume et al (2018) where not aware of the existence of equivalence tests, and we believe that our explanation of the similarities between the TOST procedure and the SGPV provides some useful context to interpret the contribution of second generation *p*-values to the statistical toolbox. The novelty lies in its use as a descriptive statistic. The added benefit of calculating the proportion of overlap of a 95% CI with the equivalence range, and using this percentage to describe the data, remains somewhat unclear for practical purposes. Nevertheless, our only goal is to clarify the relationship between a newly proposed statistic and the already existing TOST approach used to test for equivalence, and let researchers make an informed decision about which statistical approach provides the best answer to their question.

@@

Blume, J. D., McGowan, L. D., Dupont, W. D., & Greevy, R. A. (2018). Second-generation *p*-values: Improved rigor, reproducibility, & transparency in statistical analyses. PLOS ONE, 13(3), e0188299. https://doi.org/10.1371/journal.pone.0188299

Lakens, D. (2017). Equivalence Tests: A Practical Primer for t Tests, Correlations, and Meta-Analyses. Social Psychological and Personality Science, 8(4), 355–362. https://doi.org/10.1177/1948550617697177

Rogers, J. L., Howard, K. I., & Vessey, J. T. (1993). Using significance tests to evaluate equivalence between two experimental groups. Psychological Bulletin, 113(3), 553–565. http://dx.doi.org/10.1037/0033-2909.113.3.553
